Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-31 tagless/tagless-final.tex
Gone over introduction

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de-Bruijn number.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

In sum, by encoding terms using ordinary functions (the \emph{final}
rather than \emph{initial} approach), we achieve a tagless interpreter
for a typed object language, in a metalanguage with a very simple type
system \citep{hindley-principal,milner-theory}.  All the run-time errors
in \S\ref{tagproblem} (applying a nonfunction; evaluating an open term)
are now reported at compile time. Because this new interpreter uses no
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.

The interpreter above is wired directly into the
functions |b|, |lam|, and so on.  In the remainder of this paper, 
we explain how to abstract the interpreter so as
to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, and so forth.

@@

We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml \citep{metaocaml} and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.
Our code is surprisingly simple, and, in the hindsight, obvious. We
must emphasize however that direct typeful tagless interpretation of a typed
language until now has thought to be impossible in plain ML and
require extensions like GADT -- which we discuss in detail in Section
XXX.

\jacques{I quite agree that the point that needs to be revisited is
  that a \emph{direct} interpreter for a (typed) language "cannot" be
  done. see discussion in related work about the direct interpreter,
  with regards to Walid's paper. Perhaps bring some of it here?}
\jacques{I have looked at a fair bit of the literature, and I have not
  seen this done, especially not in the typed 'online' setting.  The
  one question that seems to be open is, how accurate can we make the
  PE without a tough BTA?}
Should we mention some of these points in our contributions subsection
below?

   We have realized that if we change the encoding of the term --
from data constructors to functions, we can encode a tagless
interpreter of a typed language in a typed language with no GADTs and
a very simple type system. The interpreter uses no universal type, no
pattern matching and it is patently obvious to the compiler that the
only non-termination can come from interpreting the fixpoint
combinator |fix|. If the source
language is strongly normalizing, our interpreter is total.

The term ``constructor'' functions lam, app, etc. look like free
variables in the encoding of an object term.  Defining these functions
differently gives rise to different interpreters.  Given the same term
representation but varying the interpreters, we can
	- evaluate the term to a value in the metalanguage
	- determine the size or depth of a term
    - compute a \emph{CBN} CPS transformation of a term
	- `compile' the term. This requires staging, as in MetaOCaml.
	- Surprisingly, it is also possible to write a partial evaluator!

\noindent We have implemented the above in (Meta)OCaml and Haskell.
We view our contributions as:
\begin{enumerate}
\item we interpret/compile/partially-evaluate a typed source language
   from within a typed language.  The requirements of the host type
   system are \jacques{not so onerous?  simple?},
\item a simple, clean implementation,
\item a nicely comparable Haskell/MetaOCaml version,
\item a functorial approach which deals with uniformly with all
evaluators, from interpretation to partial evaluation to CBN CPS transform,
\item clear and easy extensibility to more features,
\item only well-typed and closed terms can be evaluated, and this is 
checked statically,
\item and an approach to self-interpretation which is compatible with the
  above.  Self-interpretation turned out to be considerably harder than
  we at first expected.
\end{enumerate}

We want to reiterate that it was previously thought that insuring
that types are well-typed \emph{or} closed required advanced type
extensions, and we show that this can be done in a simple subset
of both OCaml and Haskell.

\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

\section{Base language and its tagless interpreters}\label{language}

The language we will embed is a simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

\begin{code}
  e ::= Lam hoas_fn | App e e | Fix hoas_fn |
  I int_literal | B bool_literal | Add e1 e2 | 
  Mul e1 e2 | Leq e1 e2 | IF b e-then e-else
\end{code}

This time we use higher-order abstract syntax \cite{DaleMiller}
rather than de Bruijn indices.  The
language is close to the language of \citet{xi-guarded}, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.

>>>>

\hyphenation{meta-language meta-languages}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless interpreter
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.
\ccshan{Need to cite Thiemann here and in related work?}

The interpreter above is wired directly into the
functions |b|, |lam|, |app|, and so on.  In the remainder of this paper, 
we explain how to abstract the interpreter so as
to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, and so forth.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN).
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is available%
\footnote{The URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, our OCaml evaluator obviously
terminates except if the fixpoint combinator $\mathrm{fix}$ is present
in the object program.

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that until now our achievement has been thought impossible: to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
states that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS
    transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self-interpretation compatible with the
  above.  Self-interpretation turned out to be harder than expected.
\end{enumerate}
\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

\section{The base language and its tagless interpreters}\label{language}

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to express Fibonacci, factorial, and
power.

In the sequel, we encode binding in object programs using higher-order
abstract syntax \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices.
